## Cumulative building-block build-up

To make the contribution of each mechanism explicit, the model is assembled block by block. Seven stages run the same seeded population (N = 100 agents, 90 days), with each stage adds one building block on top of the previous ones and reports the metrics that block introduces. Stage 1 is initialization only; stages 2--6 add stress perception, resilience activation, interaction, resource allocation, and stress buffering in turn; stage 7 is the full model, including the remaining daily-cycle internals (affect dynamics, PSS-10 consolidation, daily reset) and network adaptation.

```{r}
#| label: cum-setup
#| echo: false
#| message: false
#| warning: false
find_data_dir <- function() {
  candidates <- c(
    file.path(getwd(), "data", "output"),
    file.path(getwd(), "..", "data", "output"),
    file.path(dirname(getwd()), "..", "data", "output")
  )
  for (cand in candidates) {
    if (file.exists(file.path(cand, "cumulative_stages_metrics.csv"))) {
      return(normalizePath(cand))
    }
  }
  stop("data/output not found - run `pixi run cumulative-blocks` first")
}
data_dir <- find_data_dir()
```

### Stage metrics

Each stage adds one building block on top of the previous ones and the table
reports the resulting final-day population state. All seven stages share the
same seeded population and network, so differences between consecutive rows
isolate the contribution of the added block.

::: {.landscape}

```{r}
#| label: cum-stage-metrics
#| echo: false
#| results: asis
blocks <- read.csv(file.path(data_dir, "cumulative_stages_metrics.csv"),
                   check.names = FALSE)

metric_cols <- setdiff(names(blocks), c("stage", "blocks_active"))

metric_labels <- c(
  avg_pss10 = "PSS-10",
  avg_stress = "Stress",
  avg_affect = "Affect",
  avg_resilience = "Resilience",
  avg_resources = "Resources",
  stress_prevalence = "Stress prevalence",
  avg_challenge = "Challenge appraisal",
  avg_hindrance = "Hindrance appraisal",
  stress_event_rate = "Stress event rate",
  coping_success_rate = "Coping success",
  interactions_per_agent_day = "Interactions/agent-day",
  support_exchange_rate = "Support exchange rate",
  avg_regeneration = "Regeneration",
  avg_net_resource_change = "Net resource change",
  avg_buffering_strength = "Buffering strength",
  avg_stress_depletion = "Stress depletion"
)

# Transpose: metrics as rows, stages as columns
tbl <- blocks[, c("stage", metric_cols)]
rownames(tbl) <- paste0("S", tbl$stage)
tbl$stage <- NULL
tbl <- t(tbl)

# Apply human-readable labels
rownames(tbl) <- ifelse(is.na(metric_labels[rownames(tbl)]),
                        rownames(tbl),
                        unname(metric_labels[rownames(tbl)]))

# Convert to data.frame, add Metric column
tbl <- as.data.frame(tbl, stringsAsFactors = FALSE)
tbl <- cbind(Metric = rownames(tbl), tbl, row.names = NULL)

kableExtra::kable(
  tbl,
  format = "latex",
  booktabs = TRUE,
  digits = 3,
  na = "-",
  caption = "Cumulative building-block metrics by stage (final-day population state)"
) |>
  kableExtra::kable_styling(
    latex_options = c("hold_position", "scale_down"),
    font_size = 9,
    full_width = TRUE
  ) |>
  kableExtra::column_spec(
    1,
    width = "3.5cm"
  ) |>
  kableExtra::footnote(
    general = c(
      "Stage 1: Initialization.",
      "Stage 2: + Stress perception.",
      "Stage 3: + Resilience activation.",
      "Stage 4: + Interaction.",
      "Stage 5: + Resource allocation.",
      "Stage 6: + Stress buffering.",
      "Stage 7: + Full model."
    ),
    general_title = "Note:",
    threeparttable = TRUE
  )

```

:::

Each added block shifts the population metrics in a way that makes its contribution explicit. Stress perception (stage 2) updates only the appraisal dimensions and leaves the outcome states untouched, so stages 1 and 2 coincide. Resilience activation (stage 3) introduces coping, and every stress event that is coped with unsuccessfully lowers affect and resilience, raises stress, and consumes resources; no regeneration, decay, or homeostatic return opposes it yet, so the population settles into a depleted, stressed state. Interaction (stage 4) propagates affect and resilience through the network with a negativity bias stronger than the positive one, dragging both toward their lower bounds. Resource allocation (stage 5) restores the resource pool but cannot move affect or resilience, and stress buffering (stage 6) starts pushing resilience back toward its baseline, albeit with small buffering strength.

These intermediate stages are deliberately partial as they enable the stress-adding mechanisms without the restorative blocks of the daily cycle. Stage 7, the complete model, activates the remaining blocks completing the two-loop daily cycle of the production pipeline. The homeostatic pull toward baseline affect and resilience, the daily stress decay, and adaptive rewiring toward more similar peers jointly restore the equilibrium reported in the model-level tables, where affect returns near neutral, resilience near its baseline, perceived stress falls to a moderate range, and coping success rises in parallel. The phase-specific metrics (regeneration, buffering strength) are exported for stage 7 by the same model reporters that feed stages 1--6, so those cells are populated throughout; the support-exchange rates for stages 1--6 are computed from per-event interaction records, whereas the stage-7 value is the daily support-to-interaction ratio exported by the production model, so the two definitions are not directly comparable.
